On the analytic structure of the linear response susceptibility in open systems

### 2018-04-23 AnalyticStructure_main.tex
update

\documentclass[reprint,aps,prl]{revtex4-1}

@@

\subsection{Fluctuation states. A localization argument for the response density}
\lb{FluctuationState}

@@

It is possible to show that excitation eigenvalues come in pair and we focus our analysis on solutions with positive energy, \emph{i.e.} with $E>0$. Looking at the coordinate representation 
of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we observe that the Helmholtz kernel evaluated in $-E+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of $E$. 
As a consequence all the $\ket{\tilde\chi^E_p}$ states exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(E+\eps_p)$ depends from $E$ and we can 
identify the threshold value $E=|\eps_p|$ in terms of which $\ket{\phi^E_p}$ changes from a bound-state like behavior to a delocalized spherical-wave like character.  
Furthermore, the orthonormalization conditions \eqref{ExcitedStateOrthNormDef1} act as a ulterior constraint for the bound-state solutions and have the effect of determining the quantization of 
the corresponding energy levels\footnote{Sternheimer equations represent the first order expansion of a set of Hamiltonian eigenvalue equations. Consequently the properties of their solution...}.
We conclude that all the states $\ket{\tilde\chi^E_p}$ and the $\ket{\phi^E_p}$ with $E$ below the threshold value constitute a genuine \emph{discrete} set, while the ones in the unbound 
energy range behaves as generalized eigenvectors associated to a continuum spectrum.

The structure of the excitation spectrum follows from the properties of the excited states described above. To clarify this point we consider a value of $E$ that satisfies equations
\eqref{ExcitationOperatorsDef1}. In correspondence of this values equations \eqref{ExcitationOperatorsDef3} possess a solution in which \emph{at least} one of the excited states 
is not zero. We denote with $p=p_m$ the label of the $\ket{\phi^E_p}$  with the highest value of $|\eps_p|$ among the ones that contribute to \eqref{ExcitationOperatorsDef2}. From the previous 
considerations it emerge that if $E<|\eps_{p_m}|$ \emph{all} the excited states in the dyadic representation of the excitation operator possess a bound-state like behavior. Excitations 
of this type thus represent localized quantities and the associated eigenvalue are discrete. On the contrary, for $E>|\eps_{p_m}|$ some (or all) of the excited states in the dyadic representation
of the excitation behave as generalized eigenvectors and the corresponding eigenvalue belong to a continuum of accessible values.

This analysis evidences that the spectrum of the excitation operators contains two distinct sectors. The first one is composed by a (finite) set of localized excitation with associated discrete 
eigenvalues. Extending the results of \cite{boffi2016} to the equation of motion of \eqref{ExcitationOperatorsDef3} we conclude that the elements of this category behave as \emph{observables} 
quantities and the associated excitation energies have a true physical meaning, independently from the computational setup used for their representation. Conversely, the second sector is made by 
a dense set of delocalized excitations with a continuum spectrum. The elements of this set behaves as the unbound virtual orbitals analyzed in \cite{boffi2016}....

The knowledge of the excitations spectrum allows us to expand the LR susceptibility in terms of excitations. Indeed we can make usage of the spectral decomposition of the resolvent
of the Liouvillian. This procedure allows us to formally solve the Liouville equation \eqref{LiouvillianRhopomegaDef1} and we have 
\be\lb{RhopExcitationDef1}
\sop{{\cal R}}(\omega) \cdot   =
\left(\sum_E + \int\dd E   \right) \op E \,
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;.
\ee

The typical structure of the excitation spectrum is represented in figure \ref{ExcitationLandscape}. 
Excitations are separated into two categories. 
The plot evidences that below the first ionization energy $|\eps_h|$ the spectrum is purely composed by a discrete set of localized excitations. Instead, in the intermediate energy interval that extends 
from $|\eps_h|$ to the absolute value of the deepest occupied orbital, the discrete sector of the spectrum is embedded in a continuum of excitations delocalized in the whole accessible space. Lastly,
for energies higher than $|\eps_1|$ only delocalized excitations are present.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\begin{tikzpicture}[scale=0.78]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$p_h$};
\node[left] at (0,3) {$p_i$};
\node[left] at (0,1) {$p_1$};
\node[below] at (2,0) {$|\eps_h|$};
\node[below] at (4,0) {$|\eps_i|$};
\node[below] at (6,0) {$|\eps_1|$};
\node[below] at (4,-0.5) {excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape. Excitation are split in horizontal line according to the threshold levels $p_m$. Filled bullet refers to localized
excitations whereas the tick black lines describe delocalized excitation with a continuum spectrum.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

True poles vs spectral function....

\section{Applications}

Discuss the criterion for a good choice of the basis of transition. Computation of the statical polarizability in the LR formalism compared to the reference value
obtained through a finite difference method.

@@

\begin{figure}[h]
\includegraphics[scale=0.56]{c6h6_excitationLandscape.eps}
\caption{\label{c6h6_excLand}(Color online) Excitation landscape of $C_6H_6$.}
\end{figure}

>>>>

\documentclass[reprint,aps,prb]{revtex4-1}

@@

\subsection{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}

@@

It is possible to show that excitation eigenvalues come in pair and we focus our analysis on solutions with positive energy, \emph{i.e.} with $E>0$. Looking at the coordinate representation 
of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we observe that the Helmholtz kernel evaluated in $-E+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of $E$. 
As a consequence all the $\ket{\tilde\chi^E_p}$ states exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(E+\eps_p)$ depends from $E$ and we can 
identify the threshold value $E=|\eps_p|$ in terms of which $\ket{\phi^E_p}$ changes from a bound-state like behavior to a delocalized spherical-wave like character.  
Furthermore, the orthonormalization conditions \eqref{ExcitedStateOrthNormDef1} act as a ulterior constraint for the bound-state solutions and have the effect of determining the quantization of 
the corresponding energy levels \footnote{Sternheimer equations represent the first order expansion of a set of Hamiltonian eigenvalue equations. Consequently the properties of their solution...}.
We conclude that all the states $\ket{\tilde\chi^E_p}$ and the $\ket{\phi^E_p}$ with $E$ below the threshold value constitute a genuine \emph{discrete} set, while the ones in the unbound 
energy range behaves as generalized eigenvectors associated to a continuum spectrum.

The structure of the excitation spectrum follows from the properties of the excited states described above. To clarify this point we consider a value of $E$ that satisfies equations
\eqref{ExcitationOperatorsDef1}. In correspondence of this values equations \eqref{ExcitationOperatorsDef3} possess a solution in which \emph{at least} one of the excited states 
is not zero. We denote with $p=p_m$ the label of the $\ket{\phi^E_p}$  with the highest value of $|\eps_p|$, among the ones that contribute to \eqref{ExcitationOperatorsDef2}. From the previous 
considerations it emerge that if $E<|\eps_{p_m}|$ \emph{all} the excited states in the dyadic representation of the excitation operator possess a bound-state like behavior. Excitations 
of this type thus represent localized quantities and the associated eigenvalue are discrete. On the contrary, for $E>|\eps_{p_m}|$ some (or all) of the excited states in the dyadic representation
of the excitation behave as generalized eigenvectors and the corresponding eigenvalue belong to a continuum of accessible values.

% This analysis reveals that the excitations spectrum is composed by two distinct sectors. The first one is realized by the (finite) set of localized excitation with associated discrete eigenvalues. 
% Extending the results of \cite{boffi2016} to the equation of motion of \eqref{ExcitationOperatorsDef3} we conclude that the elements of this category behave as \emph{observables} 
% quantities and the associated excitation energies have a true physical meaning, independently from the computational setup used for their representation. Conversely, the second sector is made by 
% a dense set of delocalized excitations with a continuum spectrum. The elements of this set behaves as the unbound virtual orbitals analyzed in \cite{boffi2016}....

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tikzpicture}[scale=0.78]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$p_h$};
\node[left] at (0,3) {$p_i$};
\node[left] at (0,1) {$p_1$};
\node[below] at (2,0) {$|\eps_h|$};
\node[below] at (4,0) {$|\eps_i|$};
\node[below] at (6,0) {$|\eps_1|$};
\node[below] at (4,-0.5) {excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape. Excitation are split in horizontal line according to the threshold levels $p_m$. Filled bullet refers to localized
excitations whereas the tick black lines describe delocalized excitation with a continuum spectrum.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This analysis reveals that the excitations spectrum is composed by two distinct sectors. The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. Extending 
the results achieved in \cite{boffi2016} for the KS virtual orbitals, to the equations for excited states \eqref{ExcitationOperatorsDef3}, we conclude that localized excitations behave as 
\emph{observable} quantities and the associated energies possess an absolute meaning, which should not depend from the computational setup used for their representation. 
The second sector contains the dense set of delocalized excitations with an associated continuum spectrum. These excitations behave as the unbound virtual orbitals analyzed in \cite{boffi2016} and 
are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and 
the associated spectrum exhibit an explicit dependence from the size of the domain. Excitations belonging to this sector thus lose their physical meaning and have to be only understood as a tool to 
express the spectral decomposition of the Liouvillian superoperator.

The typical structure of the excitation spectrum is depicted in figure \ref{ExcitationLandscape}. The plot evidences that below the first ionization energy $|\eps_h|$ only localized excitations contribute
to the spectrum. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the absolute value of the deepest occupied orbital, the elements belonging to the discrete sector are embedded 
in a continuum of delocalized excitations. Lastly, for energies higher than $|\eps_1|$ only delocalized excitations are present.

The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
Liouvillian. Pursuing this procedure provides:
%The knowledge of the excitations spectrum allows us to expand the LR susceptibility in terms of excitations. Indeed we can make usage of the spectral decomposition of the resolvent
%of the Liouvillian. This procedure allows us to formally solve the Liouville equation \eqref{LiouvillianRhopomegaDef1} and we have 
\be\lb{RhopExcitationDef1}
\sop{{\cal R}}(\omega) \cdot   =
\left(\sum_E + \int\dd E   \right) \op E \,
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
\ee 
where we have divided the two sectors in the sum.

@@

We can express the fluctuation states in terms of the susceptibility functional
\be\lb{FluctuationStateExcitationDef1}
\ket{f_p(\omega)}   =
\left(\sum_E + \int\dd E   \right) 
\frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\op\Phi(\omega)}}{\omega-E} \ket{\phi^E_p} \;. \nn
\ee 
This relation clarifies the role of the excitations in the spectral decomposition. We know from the analysis of section \ref{FluctuationState} that fluctuation states are localized
up to $\omega=|\eps_h|$.

\section{Criterion for a reliable sampling of the excitation operators} %Applications

Discuss the criterion for a good choice of the basis of transition. Computation of the statical polarizability in the LR formalism compared to the reference value
obtained through a finite difference method. Remark the importance of using a systematic basis, only in this way the convergence of the statical polarizability turns out to 
be a meaningful criterion for the evaluation of the basis that sample the empty state space. Refer to the paper of....

@@

\begin{figure}[ht]
\includegraphics[scale=0.56]{c6h6_excitationLandscape.eps}
\caption{\label{c6h6_excLand}(Color online) Excitation landscape of $C_6H_6$.}
\end{figure}
